---
title: "Mortality EDA"
subtitle: "NurseBridge Project"
author: "Edward Yu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
      theme: readable
      highlight: tango
      toc: true
      toc_float:
        collapsed: true
      toc_depth: 4
      df_print: paged
      code_folding: hide
      fig_width: 8
      fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo    = TRUE,
                      error   = FALSE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
library(janitor)
library(magrittr)
library(readr)
library(DT)

theme_set(theme_minimal())

rm(list = ls())
```

# Intro

NurseBridge (NB) wants to create a data driven predictive model allowing rural healthcare systems to make more informed decisions regarding compensation offerings at the most granular unit of time possible.

# Data

[Goal is to use this data to categorize risk of death by day of the week on a county level.]{.underline}

[*Underlying cause of death 2018-21*](https://wonder.cdc.gov/ucd-icd10-expanded.html)

-   Data was pulled from the CDC website by grouping by County, Month, and Weekday.
-   *Suppressed* death counts were added via the CDC to protect privacy, and mask any death values number 1-9.
-   *Unreliable* was appended to calculated crude-death-rates if total deaths were below 20.

## Import

Initial import is only on Texas for the year 2021. When we determine this structure of data is sufficient for analyses further state and year data can be pulled to round off the dataset. It was mentioned that COVID-19 deaths should be excluded, so 

```{r}

make_monthly_dataset <- function(state, year, export){
  
  path1=paste0("data/", state, "_", year, "_01.csv")
  path2=paste0("data/", state, "_", year, "_02.csv")
  path3=paste0("data/", state, "_", year, "_03.csv")
  path4=paste0("data/", state, "_", year, "_04.csv")
  
  d1 <- read_csv(path1) %>% 
    clean_names() %>% 
    select(county, county_code, population, deaths) %>% 
    set_colnames(c("county", "county_code", "annual_county_population", "annual_all_deaths")) %>% 
    drop_na()
  
  d2 <- read_csv(path2) %>% 
    clean_names() %>%
    select(county_code, deaths) %>% 
    set_colnames(c("county_code", "annual_hospital_deaths")) %>% 
    drop_na() %>% 
    left_join(d1) %>% 
    select(county, county_code, annual_county_population, annual_all_deaths, annual_hospital_deaths)
  
  month_sub = paste0(year, "/")
  
  d3 <- read_csv(path3) %>% 
    clean_names() %>% 
    select(county_code, month_code, deaths) %>% 
    mutate(
      month = month(as.numeric(gsub(month_sub, "", month_code)), label=TRUE)
    ) %>% 
    select(county_code, month, deaths) %>% 
    set_colnames(c("county_code", "month", "monthly_hospital_deaths")) %>% 
    drop_na() %>% 
    right_join(d2) %>% 
    select(county, county_code, annual_county_population, annual_all_deaths, annual_hospital_deaths, month, monthly_hospital_deaths)
  
  d4 <- read_csv(path4) %>% 
    clean_names() %>% 
    select(county_code, month_code, deaths) %>% 
    mutate(
      month = month(as.numeric(gsub(month_sub, "", month_code)), label=TRUE)
    ) %>% 
    select(county_code, month, deaths) %>% 
    set_colnames(c("county_code", "month", "monthly_covid_hospital_deaths")) %>% 
    drop_na() %>% 
    right_join(d3) %>% 
    mutate(year = as.numeric(year),
           state = as.factor(state),
           county = as.factor(gsub(" County, TX", "", county))
    ) %>% 
    select(year, state, county, county_code, annual_county_population, annual_all_deaths, annual_hospital_deaths, month, monthly_hospital_deaths, monthly_covid_hospital_deaths)
  
  if(export){
    export_filename = paste0(state, "_", year, "_monthly")
    write_csv(d4, paste0("export/", export_filename, ".csv"))
  }
  
  return(d4)
}

make_wkday_dataset <- function(state, year, export){
  
  path5=paste0("data/", state, "_", year, "_05.csv")
  path6=paste0("data/", state, "_", year, "_06.csv")
  
  month_sub = paste0(year, "/")

  d5 <- read_csv(path5) %>% 
    clean_names() %>% 
    mutate(
      month = month(as.numeric(gsub(month_sub, "", month_code)), label=TRUE)
    ) %>% 
    select(county_code, month, weekday, deaths) %>% 
    set_colnames(c("county_code", "month", "weekday", "wkday_hospital_deaths")) %>% 
    drop_na()
  
  d6 <- read_csv(path6) %>% 
    clean_names() %>% 
    mutate(
      month = month(as.numeric(gsub(month_sub, "", month_code)), label=TRUE)
    ) %>% 
    select(county_code, month, weekday, deaths) %>% 
    set_colnames(c("county_code", "month", "weekday", "covid_wkday_hospital_deaths")) %>% 
    drop_na()
  
  d6 <- d5 %>% left_join(d6)
  
  if(export){
    export_filename = paste0(state, "_", year, "_wkday")
    write_csv(d6, paste0("export/", export_filename, ".csv"))
  }
 
  return(d6)
   
}

TX_2021_monthly <- make_monthly_dataset(state = "TX", year = "2021", export = FALSE)
TX_2021_wkday <- make_wkday_dataset(state = "TX", year = "2021", export = FALSE)

TX_2021_monthly %>% glimpse()
TX_2021_wkday %>% glimpse()
```

## Notes {.tabset .pills}

Shortcomings become more evident the more narrowly we focus our data. Suppressed data becomes more and more abundant as we drill down into tallying deaths by weekdays. The effect would be exacerbated if skimming off COVID-related deaths, possibly making a large portion of the data unusable.

### Monthly data

Most data is visible on the monthly table. We start losing data due to `Suppression` when we get to Counties with smaller populations

```{r}
DT::datatable(TX_2021_monthly)
```

### Weekday data
```{r}
DT::datatable(TX_2021_wkday)
```



<!-- # EDA -->

<!-- ## Total deaths and population -->

<!-- How have total deaths in the state trended over the years compared to population? -->

<!-- ```{r} -->
<!-- TX_all %>% -->
<!--   select(county, year, annual_county_population) %>%  -->
<!--   distinct() %>%  -->
<!--   group_by(year) %>%  -->
<!--   summarise(TX_population = sum(annual_county_population)) %>%  -->
<!--   mutate(year = as.factor(year)) %>%  -->
<!--   left_join( -->
<!--     TX_all %>% -->
<!--       select(county, year, annual_county_deaths) %>%  -->
<!--       distinct() %>%  -->
<!--       group_by(year) %>%  -->
<!--       summarise(total_deaths = sum(as.numeric(annual_county_deaths), na.rm = TRUE)) %>%  -->
<!--       mutate(year = as.factor(year)) -->
<!--   ) %>%  -->
<!--   pivot_longer( -->
<!--     cols = TX_population:total_deaths) %>%  -->
<!--   ggplot(aes(x=year, y=value, group=name, fill=name)) + -->
<!--   # geom_col(position="dodge") -->
<!--   geom_point()+ -->
<!--   geom_line()+ -->
<!--   facet_wrap(~name, scales="free_y") +  -->
<!--   scale_y_continuous(labels=comma) -->
<!-- ``` -->

<!-- ## Annual patterns -->

<!-- Do certain months show patterns year after year? -->

<!-- ```{r} -->
<!-- gg_1 <-  -->
<!--   TX_all %>%  -->
<!--     mutate_if(is.character, as.numeric) %>%  -->
<!--     mutate(year = as.factor(year)) %>%  -->
<!--     group_by(year, month) %>%  -->
<!--     summarise(deaths = sum(daily_county_deaths, na.rm = TRUE)) %>%  -->
<!--     # ungroup() %>%  -->
<!--     ggplot(aes(x=month, y=deaths, color=year, group=year)) + -->
<!--     geom_point() +  -->
<!--     geom_line(alpha=.8) +  -->
<!--     labs( -->
<!--       title = "Deaths over time" -->
<!--     ) +  -->
<!--     scale_y_continuous(labels = comma) +  -->
<!--   xlab("Month") + -->
<!--   ylab("Deaths") -->

<!-- ggplotly(gg_1) -->
<!-- ``` -->

<!-- ## Annual adjusted -->

<!-- Every year the deaths increase, what about after normalizing against population? -->

<!-- ```{r} -->
<!-- # get TX population by year -->
<!-- TX_pop <- TX_all %>% -->
<!--   select(county, year, annual_county_population) %>%  -->
<!--   distinct() %>%  -->
<!--   group_by(year) %>%  -->
<!--   summarise(TX_population = sum(annual_county_population)) %>%  -->
<!--   mutate(year = as.factor(year)) -->

<!-- gg_2 <- TX_all %>%  -->
<!--   mutate_if(is.character, as.numeric) %>%  -->
<!--   mutate(year = as.factor(year)) %>%  -->
<!--   group_by(year, month) %>%  -->
<!--   summarise(deaths = sum(daily_county_deaths, na.rm = TRUE)) %>%  -->
<!--   left_join(TX_pop) %>% -->
<!--   mutate(pct_pop_deaths = deaths/TX_population) %>%  -->
<!--   ggplot(aes(x=month, y=pct_pop_deaths, color=year, group=year)) + -->
<!--   geom_point() +  -->
<!--   geom_line(alpha=.8) +  -->
<!--   labs( -->
<!--     title = "Deaths over time", -->
<!--     subtitle = "Normalized for population" -->
<!--   ) +  -->
<!--   scale_y_continuous(labels = percent) +  -->
<!--   xlab("Month") + -->
<!--   ylab("Deaths (percent of state population") -->
<!-- ggplotly(gg_2) -->

<!-- ``` -->

<!-- ## Covid spike {.tabset .tabset-fade .tabset-pills} -->

<!-- ### Standard -->

<!-- Are the spikes seasonal or due to covid? -->

<!-- ```{r} -->

<!-- ## visualize covid spike ---- -->
<!-- gg_3 <- TX_all %>%  -->
<!--   mutate_if(is.character, as.numeric) %>%  -->
<!--   mutate(year = as.factor(year)) %>% -->
<!--   group_by(year, month) %>%  -->
<!--   summarise(deaths = sum(daily_county_deaths, na.rm = TRUE)) %>%  -->
<!--   left_join(TX_pop) %>% -->
<!--   mutate( -->
<!--     pct_pop_deaths = deaths/TX_population, -->
<!--     date = ymd(paste0(year,"-", month, "-01")) -->
<!--     ) %>%  -->
<!--   ggplot(aes(x=date, y=pct_pop_deaths, color=year)) + -->
<!--   geom_point()+ -->
<!--   geom_line(alpha=.8)+ -->
<!--   labs( -->
<!--     title = "Deaths over time", -->
<!--     subtitle = "Normalized for population" -->
<!--   ) +  -->
<!--   scale_y_continuous(labels = percent) +  -->
<!--   scale_x_date(date_breaks = "4 months", date_labels = "%Y-%m") +  -->
<!--   xlab("Month") + -->
<!--   ylab("Deaths (percent of state population")+ -->
<!--   geom_rect(aes(xmin=ymd("2020-01-01"), xmax=ymd("2021-12-01"), ymin=0, ymax=.001), alpha=.005, fill='orange', color=NA) + -->
<!--   theme(legend.position = "none")   -->
<!-- gg_3 -->

<!-- ``` -->

<!-- ### Plotly -->

<!-- ```{r} -->
<!-- ggplotly(gg_3) -->
<!-- ``` -->

<!-- ## Monthly patterns -->

<!-- If we look at how deaths have trended over time, on a per month basis, can we see any useful patterns at all? -->

<!-- ```{r} -->

<!-- TX_all %>%  -->
<!--   mutate_if(is.character, as.numeric) %>%  -->
<!--   mutate(year = as.factor(year)) %>% -->
<!--   group_by(year, month) %>%  -->
<!--   summarise(deaths = sum(daily_county_deaths, na.rm = TRUE)) %>%  -->
<!--   left_join(TX_pop) %>% -->
<!--   mutate( -->
<!--     pct_pop_deaths = deaths/TX_population, -->
<!--     date = ymd(paste0(year,"-", month, "-01")) -->
<!--   ) %>%  -->
<!--   ggplot(aes(x=year, y=deaths))+ -->
<!--   geom_point()+ -->
<!--   geom_line(group=1,alpha=.4)+ -->
<!--   facet_wrap(~month) -->

<!-- ``` -->

<!-- ## Weekday -->

<!-- Are certain weekdays more deadly? -->

<!-- ```{r} -->
<!-- TX_all %>%  -->
<!--   mutate_if(is.character, as.numeric) %>%  -->
<!--   select(county, year, month, weekday, daily_county_deaths) %>%  -->
<!--   group_by(weekday) %>%  -->
<!--   summarise(total_deaths = sum(daily_county_deaths, na.rm = TRUE)) %>%  -->
<!--   ggplot(aes(x=weekday, y=total_deaths))+ -->
<!--   geom_point()+ -->
<!--   geom_line(group=1)+ -->
<!--   scale_y_continuous(labels=comma) -->
<!-- ``` -->

<!-- And does this pattern hold true year by year? -->

<!-- ```{r} -->
<!-- TX_all %>%  -->
<!--   mutate_if(is.character, as.numeric) %>%  -->
<!--   select(county, year, month, weekday, daily_county_deaths) %>%  -->
<!--   group_by(year, weekday) %>%  -->
<!--   summarise(total_deaths = sum(daily_county_deaths, na.rm = TRUE)) %>%  -->
<!--   ggplot(aes(x=weekday, y=total_deaths))+ -->
<!--   geom_point()+ -->
<!--   geom_line(group=1)+ -->
<!--   scale_y_continuous(labels=comma)+ -->
<!--   facet_wrap(~year, scale="free_y", ncol=1) -->
<!-- ``` -->



<!-- ## -->

<!-- * Filter out COVID `U07.1 (COVID-19)` -->
<!-- * Maybe filter for in hospital -->
<!-- * Send data in `County | Month, year | Deaths` format -->
<!-- * Send data in `County | Weekday, Month, year | Deaths` format -->
