---
title: "Mortality EDA"
subtitle: "NurseBridge Project"
author: "Edward Yu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
      theme: readable
      highlight: tango
      toc: true
      toc_float:
        collapsed: true
      toc_depth: 4
      df_print: paged
      code_folding: hide
      fig_width: 8
      fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo    = TRUE,
                      error   = FALSE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
library(janitor)
library(magrittr)
library(readr)
library(DT)
library(plotly)

theme_set(theme_minimal())

rm(list = ls())

df_wkday   <- read_csv("export2/df_wkday.csv")   %>% mutate_if(is.character, as.factor)
df_monthly <- read_csv("export2/df_monthly.csv") %>% mutate_if(is.character, as.factor)
df_annual  <- read_csv("export2/df_annual.csv")  %>% mutate_if(is.character, as.factor)
```

# Intro

NurseBridge (NB) wants to create a data driven predictive model allowing rural healthcare systems to make more informed decisions regarding compensation offerings at the most granular unit of time possible.

# Data

[Goal is to use this data to categorize risk of death by day of the week on a county level.]{.underline}

We will start with monthly data and attempt to work into the day of the week. Due to anonymization of data, days of the week are specified but not the actual dates. I.e, All Fridays of the month are bundled into a single Friday variable. 

[*Underlying cause of death 2018-21*](https://wonder.cdc.gov/ucd-icd10-expanded.html)

-   Data was pulled from the CDC website by grouping by County, Month, and Weekday.
-   *Suppressed* death counts were added via the CDC to protect privacy, and mask any death values number 1-9.
    * Values replaced with `5`.
-   *Unreliable* was appended to calculated crude-death-rates if total deaths were below 20.

## Glimpse

Initial imports were performed on Texas and Oklahoma ranging from 2018-21 as these two locations were specifically mentioned by NB as some of their primary areas of focus.  

Should be noted that when deaths are between 1-9, they are displayed as "Suppressed", to preserve privacy.

```{r}


```



